---
title: "R_Data_Visualisation"
author: "Amdouni_Youssef"
output: html_document
---

# Data Visualisation with ggplot2 & plotly 
#### ggplot2 is the most popular R package for data visualization, it is built on the idea of adding layers "Grammer of Graphics"
* Data: dataframe we want to explore 
* Aesthetics: specify features we want to plot 
* Geometries: function to represent data points (histogram, barplot, ...) 
* Facets: allow us to put multiple plots on a single canvas  
* Statistics: a statistical transformation performs some useful statistical summary
* Coordinates: deals with plot axis  
* Theme: add a theme to our plot to affect things such as coloring background

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(plotly)
```


```{r}
head(mtcars)
```
```{r}
str(mtcars)
```

```{r}
# Data + Aesthetics + Geometries 
ggplot(data = mtcars, aes(x=mpg, y=hp)) + geom_point()
```


```{r message=FALSE, warning=FALSE}
# Data + Aesthetics + Geometries + Facets + Statistics + Coordinates + Theme
ggplot(data = mtcars, aes(x=mpg, y=hp)) + geom_point() + facet_wrap(cyl ~ .) + stat_smooth() + coord_cartesian(xlim = c(10, 35)) + xlab("Miles per Gallon") + ylab("Gross horsepower") + ggtitle("Gross horsepower function of miles per gallon") + theme_dark() + theme(plot.title = element_text(hjust=0.5, color = "blue", face = "bold", size = 14))
```
```{r message=FALSE, warning=FALSE}
ggplot(mtcars, aes(mpg, wt, colour = factor(cyl))) + geom_point(aes(size=wt), alpha=0.7) + xlab("Miles per Gallon") + ylab("Gross horsepower") + theme_bw()
```
```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) + geom_boxplot() + ylab("Miles/Gallon") + xlab("Number of cylinders") + theme_bw()
```
Now we will use the mpg data to explore other Titanic geometries 
```{r message=FALSE, warning=FALSE}
data <- read.csv("data/titanic_train.csv")
head(data)
```
```{r message=FALSE, warning=FALSE}
ggplot(data = data, aes(x = Age)) + geom_histogram(binwidth = 1, color = "black", fill = "pink", alpha = 0.8) +   labs(title = "Age distribution in the Titanic data", x = "Age", y = "Number of passengers") + theme_bw() +  theme(plot.title = element_text(size = 14, color = "red", hjust = 0.5))
```


```{r message=FALSE, warning=FALSE}
ggplot(data = data, aes(x = Age, fill = factor(Survived))) + geom_histogram(binwidth = 1, position = "identity", alpha = 0.5) +   labs(title = "Age distribution in the Titanic data", x = "Age", y = "Number of passengers") + theme_bw() +  theme(plot.title = element_text(size = 14, color = "blue", hjust = 0.5))

```
```{r}
ggplot(data = data, aes(x = factor(Survived))) + geom_bar(aes(fill = factor(Pclass))) + theme_bw()
```
```{r}
ggplot(data = data, aes(x = factor(Survived), fill = factor(Pclass))) + geom_bar(position = position_dodge(width = 1)) +
labs(title = "Survival by Pclass", x = "Survived") + theme_bw() + theme(plot.title = element_text(hjust = 0.5, size = 15)) 
```