---
title: "RBasics2"
author: "Amdouni_Youssef"
output: html_document
---
### R Lists 

list allow us to store a variety of data structures (vector,matrix, data frame, etc.) under a single variable.
```{r}
newList <- list(seq(1, 11, by = 2), list("Hello world", "R"))
newList
str(newList)
```
```{r}
# extract element as vector 
newList[[1]] 
newList[[2]][[1]]
newList[[2]][[2]]
# extract item as a list
newList[[2]][1]

# extract the numbers item as a list
newList[1]

# multiply each element in the numbers item by 2
newList[[1]] <- newList[[1]] * 2
newList[[1]]
```
```{r}
# name the items in the list 
names(newList) <- c("Num", "Char")
newList
```
```{r}
newList$Num
newList$Num <- newList$Num / 2
newList$Num
newList$Letter <- c("a", "b", "c")
str(newList)
```
```{r}
# remove "the iPhone"b" from the Letter item
#newList[[3]] <- newList[[3]][-2]
newList$Letter <- newList$Letter[-2]
newList

# remove Letter item from the list
newList$Letter <- NULL
# newList[[3]] <- NULL
newList
```
### Function with R
```{r}
flip <- function(){
  coin <- c("heads", "tails")
  toss <- sample(coin, 10, replace = TRUE, prob = c(0.5, 0.5))
  return(toss)
}
flip()
```
### For loops
```{r}
sum = 0
for(i in 1:6){
  sum <- sum + i
  print(sum)
} 
```
### while loops
```{r}
sum = 2
iter = 1
while(sum<8){
  sum = sum + iter 
  print(sum)
}
```
### Logical operators

- AND - &
- OR - |
- NOT - !

Difference between | , || , & and &&

- Comparison of the entire vector elements by element => | , &
- Comparison of the first elements in the vectors => || , &&
```{r}
(-2:2) >= -1
(-2:2) <= 1
((-2:2) >= -1) & ((-2:2) <= 1)
((-2:2) >= -1) | ((-2:2) <= 1)
((-2:2) >= -1) && ((-2:2) <= 1)
((-2:2) >= -1) || ((-2:2) <= 1)

```
### Conditions with R
```{r}
set.seed(128)
mark <- runif(1, min = 2, max = 20)
if (mark >= 18){
  print("Excelent")
}else if (mark>=15 & mark<18){
  print("good")
}else if (mark<15 & mark>=10){
  print("Fair")
}else{
  print("Poor")
}
```
## Build-IN R functions 
| Functions     | Description   |
|---------------|:-------------:|
| abs(x)        | absolute value | 
| sqrt(x)       | square root      |   
| min(x), max(x)| min, max      |    
| log(x)	      | natural logarithm |
| log10(x)	    | common logarithm |
| exp(x)        | exponential |
| cos(x), acos(x), acosh(x), etc.| cosinus, etc.|
| ceiling(x)    | returns the smallest integer not less than x; ceiling(3.3) = 4 |
| floor(x)      | returns the largest integer not greater than x; floor(3.9) = 3 |
| round(x, digit = n) | rounds the values in its first argument; round(3.456, 2) = 3.46 |
| sort(x) | sort a vector or a factor; sort(c(5,2,9,1)) = 1 2 5 9 |
| rev(x) | reverse vector items; rev(c(5,2,9,1)) = 1 9 2 5 |
| append(x1, x2) | append item to vector; append(c(1,2,3), c(4,5)) = 1 2 3 4 5 |

### Apply functions 
Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix. There many apply function such as:

- `lapply(X, FUN, ..)`: returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X. 
- `lapply(X, FUN, ..)`: returns a vector of the same length as X, each element of which is the result of applying FUN to the corresponding element of X. 

```{r}
vector <- seq(from = 2, to = 11, by = 2)
# apply squared root of each element in vector
lapply(vector, sqrt)
```
```{r}
sapply(vector, sqrt)
```
```{r}
sapply(vector, function(x){x^2})
```
### Timestamps
#### Date
To covert string to Date object, it must be in the standard time format "year-month-day". 

| code     | Value   |
|---------------|:-------------:|
| %d      | Day of the month (decimal number) | 
| %m      | Month (decimal number) |   
| %b      | Month (abbreviated)    |    
| %B	    | Month (full name) |
| %y	    | Year (2 digit) |
| %Y      | Year (4 digit) |

```{r}
# Sys.Date() To get today's day
as.Date('1920-10-04')
```
```{r}
as.Date("october/4/1920", format="%B/%d/%Y")
```

```{r}
as.Date("Oct/4/1920", format="%b/%d/%Y")
```
#### Time
For more information `help(strptime)`
```{r}
as.POSIXct("october/4/1920 09:30:02", format="%B/%d/%Y %H:%M:%S")
```

